---
title: "About Me"
---

```{r, echo = FALSE}
library(knitr)
"print.magick-image" <- function(x, ...){
  ext <- ifelse(length(x), tolower(image_info(x[1])$format), "gif")
  tmp <- tempfile(fileext = paste0(".", ext))
  image_write(x, path = tmp)
  knitr::include_graphics(tmp)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### TL;DR
I'm a marine biologist and science communicator by training, but I'm taking all of the data analysis, statistics, and mathematical modeling skills I've learned and I'm now applying them to new problems.

### The Long Version
I've always loved science, exploration, and discovery.  My natural curiosity, love of animals and the environment, and constant pursuit of knowledge led me into a marine biology and chemistry double major in college (BS/BA) and a Master's degree in Marine Sciences. That generally leads people to think that I was "playing with animals" all day. 

And while I *did* get to work with some amazing animals over the years...

```{r,  echo = FALSE, message = FALSE, fig.align = "center"}
library(magick)
library(magrittr)


turtle <- image_read("images/turtle.jpg") %>% 
  image_scale("400") %>% 
  image_scale("x400") %>% 
  image_border("white", "10x10")

dolphin <- image_read("images/dolphin.jpg") %>% 
  image_scale("400") %>% 
  image_scale("x400") %>% 
  image_border("white", "10x10")

seal <- image_read("images/seal_1.jpg") %>% 
  image_scale("400") %>% 
  image_scale("x400") %>% 
  image_border("white", "10x10")

animals <- c(turtle, dolphin, seal)


all_animals <- image_append(animals)
print(all_animals)

```




